---
title: "EDA-Report-JALT"
subtitle: "STATX290 Statistical Practice Exploratory Data Analysis Report"
author: "Team JALT - Jerry, Aysha, Lily, and Theo"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    embed-resources: false
    theme: cosmo
execute:
  warning: false
  message: false
  echo: true
---

# Introduction

This EDA report looks at polling data from 2012 and 2016 US pre-election polling. 

The data was obtained from HuffPost Pollster at https://projects.huffingtonpost.com. There are three data sets:

- `state_polls_2012.csv`
- `state_polls_2016.csv`
- `1976-2024-president.csv`

In the raw data sets, each row represents a separate poll containing the candidates for that year, the polling dates, 

An additional data set, `1976-2024-president.csv`, along with the file `codebook-us-president-1976-2020.md` is included.

# Description of Data Cleaning

Setting up libraries and reading in data.

```{r}
#| label: setup

# Installing packages for cleaning ----------------------------------------

library(tidyverse)   # dplyr, ggplot2, tidyr, readr, stringr, forcats
library(here)        # project-relative paths; don't use setwd()
library(lubridate)   # dates
library(glue)        # easy string manipulation

# Set up directories ------------------------------------------------------

raw_dir   <- here("data", "raw")
clean_dir <- here("data", "clean")

# Reading in raw data -----------------------------------------------------

polls_2016_raw <- read_csv(glue("{raw_dir}/state_polls_2016.csv"))
polls_2012_raw <- read_csv(glue("{raw_dir}/state_polls_2012.csv"))
results_1976_2024_raw <- read_csv(glue("{raw_dir}/1976-2024-president.csv"))

```


Extracting the time to election and creating a new variable.

```{r}
#| label: clean-data-dates

# Cleaning the datasets ---------------------------------------------------

# Time to election
# Define election day = first Tuesday after the first Monday in November.
election_day <- c("2012" = as.Date("2012-11-06"),
                  "2016" = as.Date("2016-11-08"))

# Adding time to election as column to data sets
polls_2016 <- polls_2016_raw |>
  mutate(
    start_date = as.Date(start_date),
    end_date   = as.Date(end_date),
    days_to_election = as.integer(election_day["2016"] - end_date)
  )
polls_2012 <- polls_2012_raw |>
  mutate(
    start_date = as.Date(start_date),
    end_date   = as.Date(end_date),
    days_to_election = as.integer(election_day["2012"] - end_date)
  )
```


The `poll_info` is muddled. From it, extract the state and year and create new variables.

```{r}
#| label: clean-data-extract-year-state
# Extracting the year and state name for each row
polls_2016 <- polls_2016 |>
  mutate(
    year = str_extract(poll_info, "^\\d{4}"),
    state = str_extract(poll_info, "(?<=^\\d{4}-).*?(?=-president)")
  )
polls_2012 <- polls_2012 |>
  mutate(
    year = str_extract(poll_info, "^\\d{4}"),
    state = str_extract(poll_info, "(?<=^\\d{4}-).*?(?=-president)")
  )
```


Rather than having each candidate as a separate variable, create a new row for each candidate in each poll.

The sample size is entered as "-1" for 2 polls, one from 2012 and one from 2016. These should be NAs, because the "-1" is being treated as an actual data point and is skewing the mean slightly negatively.

```{r}
#| label: clean-data-long

# Creating longer data sets with each candidate on a new row
candidate_cols_2016 <- c("Trump", "Clinton", "Johnson", "McMullin", 
                         "Other", "Undecided")
polls_2016_long <- polls_2016 |>
  pivot_longer(cols = all_of(candidate_cols_2016),
               values_transform = as.numeric,
               names_to = "candidate", values_to = "pct")

candidate_cols_2012 <- c("Obama", "Romney", "Other", "Undecided")
polls_2012_long <- polls_2012 |>
  pivot_longer(cols = all_of(candidate_cols_2012),
               values_transform = as.numeric,
               names_to = "candidate", values_to = "pct")

# Combining datasets into long set ----------------------------------------
# Removing negative population values
polls_long <- bind_rows(polls_2012_long, polls_2016_long) |>
  mutate(sample_size = na_if(sample_size, -1),
         across(c(sample_subpopulation, mode, partisanship,
                  partisan_affiliation, state, candidate),
                factor)
  )

# Saving data sets as csv in clean folder ---------------------------------

write_csv(polls_long, "data/clean/polls_long_clean.csv")    # combined date
write_csv(polls_2012, "data/clean/polls_2012_clean.csv")    # clean 2012 data
write_csv(polls_2016, "data/clean/polls_2016_clean.csv")    # clean 2016 data
write_csv(polls_2012_long, "data/clean/polls_2012_long_clean.csv")  # 2012 long
write_csv(polls_2016_long, "data/clean/polls_2016_long_clean.csv")  # 2016 long
```

- What does this data represent?
- What is measured?
- How was it collected?

*Is there any data missingness or other errors? How can this be corrected?*

- Do the measurements make sense in context? (you might need to discuss with the instructor what to do if not)




# Exploratory Data Analysis

```{r}
#| label: time-plot
#| fig-cap: "Temporal distribution of polling effort, 2016."
polls_2016 |>
  ggplot(aes(days_to_election)) +
  geom_histogram(binwidth = 7) +
  scale_x_reverse() +
  labs(x = "Days to election (binned by week)", y = "Number of polls")

#| label: time-plot
#| fig-cap: "Temporal distribution of polling effort, 2012."
polls_2012 |>
  ggplot(aes(days_to_election)) +
  geom_histogram(binwidth = 7) +
  scale_x_reverse() +
  labs(x = "Days to election (binned by week)", y = "Number of polls")
```


(Use summaries and visualisations to identify relationships in the data, and motivate your research questions)

*What relationships can you observe visually?*
Here, maybe include our graphs from week 3 forum?



```{r}
#| label: Candidates 2012 v Percentage (pre-polling)

polls_2012_long |> filter(year %in% c("2012")) |>
  ggplot() + 
  geom_boxplot(aes(x = candidate, y = pct, fill = candidate)) +
  labs(x = "Candidates 2012", y = "Percentage") + 
  theme_bw()
```

Discussion: From this data set we could explore how accurate the pre-polling data was to the outcome data from the results data set. This may need to be changed because the American voting system depends on the state and each state had different amount of polls which means states that has less polls or was not included will not be represented. This will bias the pre-polling towards highly polled states.

```{r}
#| label: plot-obama-votes-over-time

polls_2012 |> ggplot(aes(y = Obama,
                         x = as.integer(difftime(ymd("2012-11-06"), end_date, units = "days")))
) + 
  geom_point(color = "darkcyan", shape = 1, stroke = 1, alpha = 0.6) +
  labs(x = "Time until election (days)",
       y = "Percent voted for Obama",
       title = str_wrap("Relationship between percentage of people 
       who voted for Obama and time until election day")) +
  scale_x_reverse() +
  scale_y_continuous(limits = c(0, NA))
```

Discussion: The purpose of this graph to to see how pre-polling votes change over time for a Obama in 2012. Just by looking at the data there does not seem to be much change in opinions over time. The main aspect to see is that more polls were completed closer to the election date. We may be able to find a better trend with more analysis. Exploring voting opinions over time may be limited by the lack of data for more than 300 days before election. Depending what we are trying to explore we may not have not enough data.

```{r}
#| label: plot-partisan-affilitation-votes-for-obama

polls_2012_long |> filter(candidate %in% c('Obama')) |>
  ggplot() +
  geom_boxplot(aes(x = partisan_affiliation, y = pct, fill = partisan_affiliation)) +
  labs(x = "Partisan Affilitation", y = "Percentage of Poll for Obama") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_viridis_d()
```

Discussion

```{r}
#| label: plot-voting-methods-for-romney

ggplot(polls_2012, aes(x = mode, y = Romney, fill = mode)) +
  geom_boxplot() +
  labs(x = "Method of Voting", y = "Percentage of Votes for Romney") +
  theme_bw() +
  theme(legend.position = "none")
```


```{r}
#| label: plot-results-2016-percentage-of-state-votes
results_1976_2024_raw |> filter(year %in% (2016)) |>
  ggplot(aes(x = party_simplified, y = (candidatevotes/totalvotes), fill = party_simplified)) +
  geom_boxplot() +
  labs(x = "Party", y = "Percentage of Votes by State") +
  theme_bw() +
  theme(legend.position = "none")
```
```{r}
#| label: plot-results-party-results-over-time
results_1976_2024_raw |>
  ggplot(aes(x = as.numeric(year), y = (candidatevotes/totalvotes), fill = party_simplified)) +
  geom_col() +
  labs(x = "Year", y = "Percentage of Votes by State", fill = "Party") +
  theme_bw() +
  theme(legend.position = "bottom")
```


```{r}
#| label: col-graph-results-2016-percentage-votesin each state
results_1976_2024_raw |> filter(year %in% (2016)) |> 
  ggplot(aes(x = state_po, y = (candidatevotes/totalvotes), fill = party_simplified)) +
  geom_col() +
  labs(x = "State", y = "Percentage of Votes") +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r}
#| label: col-graph-polling-2016-percentage-in each state
polls_2016_long |>
  ggplot(aes(x = state, y = mean(pct, group_by((state))), fill = candidate)) +
  geom_col() +
  labs(x = "State", y = "Percentage of poll") +
  theme_bw() +
  theme(legend.position = "bottom")


polls_2016_long |> group_by((state)) |>
  mean(pct, na.rm = TRUE)

```


# Research Questions

- How representative were opinion polls of the election results?