---
title: "EDA-Report-JALT"
subtitle: "STATX290 Statistical Practice Exploratory Data Analysis Report"
author: "Team JALT - Jierui, Aysha, Lily, and Theo"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    embed-resources: false
    theme: cosmo
execute:
  warning: false
  message: false
  echo: true
---

# Introduction

This EDA report looks at polling data from 2012 and 2016 US pre-election polling. 

The data was obtained from HuffPost Pollster at [https://projects.huffingtonpost.com](https://projects.huffingtonpost.com). There are three datasets:

- `state_polls_2012.csv`
- `state_polls_2016.csv`
- `1976-2024-president.csv`

In the raw datasets `state_polls_2012.csv` and `state_polls_2016.csv`, each row represents a separate poll containing the candidates for that year, poll ID, pollster, the polling start and end dates, sample sub-population and size, polling mode (method), partisanship, and partisan affiliation.

The dataset `1976-2024-president.csv` contains the state-level results of elections from 1976 to 2024. The included file `codebook-us-president-1976-2020.md` provides detailed descriptions of each variable in this dataset. It is important to note that the codebook is not up to date, as the title date range does not match the data range. 

NOTES
## Discuss
What population is captured in a **poll** row versus a **results** row? Who may be *excluded*
from each? For the polling datasets, distinguish the target population (who should be included in the survey) from the sampling frame (who is actually surveyed).
:::

*Your answer:*
Poll rows: state, Candidates, poll_id, pollster, start/end date, sample sub [population, mode, sample size, partisanship, affiliation
Results: Year, state, candidate who won in place, party details, candidate votes, total votes 

The candidate columns in the polls are reported **percentages, not counts**.
Who chose the denominator? Are they percentages of all respondents, or of
decided voters only? How would you tell?
:::

*Your answer:*
They are obviously percentages because the rows add up to (roughly, given rounding) 100, and don't correspond to any total number.
All voters, decided and undecided, are included in the percentage (that's how percentages work).

The columns `sample_subpopulation`, `mode`, `partisanship`, and `partisan_affiliation` describe characteristics of the polls. How much variation is in these columns? Are there any potential problems for statistical inference you should note?
:::

*Your answer:*
It depends if it represents the population of voting Americans accurately as a good sample. 
The majority of those polled had no affiliation, and there were roughly twice as many Dems and Reps (~200 v ~100)

## Discuss
Polls are weighted and adjusted by each pollster using (mostly) undisclosed procedures.
What does the polling dataset tell you about the polling procedures? What decisions/rules were likely to be undisclosed?
How might this affect subsequent analysis?
:::

*Your answer:* They rounded the percentages so they no longer add nicely to 100%. Undisclosed decision about if people don't answer/ don't answer properly. Unclear if all polls had the option to choose other or undecided.

# Description of Data Cleaning

Set up the required libraries and read in the raw data.

```{r}
#| label: setup-libraries
#| echo: FALSE

library(tidyverse)   # dplyr, ggplot2, tidyr, readr, stringr, forcats
library(here)        # project-relative paths; don't use setwd()
library(lubridate)   # dates
library(glue)        # easy string manipulation
library(janitor)     # for the clean_names() function
```


```{r}
#| label: setup-read-data

# Set up directories
raw_dir   <- here("data", "raw")
clean_dir <- here("data", "clean")

# Read in raw data
polls_2016_raw <- read_csv(glue("{raw_dir}/state_polls_2016.csv"))
polls_2012_raw <- read_csv(glue("{raw_dir}/state_polls_2012.csv"))
results_1976_2024_raw <- read_csv(glue("{raw_dir}/1976-2024-president.csv"))

# Read in external data
income_raw <- read_csv(glue("{raw_dir}/income_by_state.csv"), skip = 3, col_names = TRUE)
poverty_raw <- read_csv(glue("{raw_dir}/poverty_by_state.csv"), skip = 3, col_names = TRUE)
unemployment_raw <- read_csv(glue("{raw_dir}/unemployment_by_state.csv"), skip = 3, col_names = TRUE)
smokefree_rate_raw <- read_csv(glue("{raw_dir}/smokefree_rate_by_state.csv"), skip = 4, col_names = TRUE)
education_raw <- read_csv(glue("{raw_dir}/education_by_state.csv"), skip = 4, col_names = TRUE)

```

The raw data needs cleaning to make it workable. We need to alter existing variables, add additional variables, and manage missingness.

The **time recording** in the polling data is arbitrary. It is more informative define a variable that counts down to the election date, as follows.

```{r}
#| label: clean-data-days-to-election

# Define election day = first Tuesday after the first Monday in November.
election_day <- c("2012" = as.Date("2012-11-06"),
                  "2016" = as.Date("2016-11-08"))

# Add time to election as column to datasets
polls_2016 <- polls_2016_raw |>
  mutate(
    start_date = as.Date(start_date),
    end_date   = as.Date(end_date),
    days_to_election = as.integer(election_day["2016"] - end_date)
  )

polls_2012 <- polls_2012_raw |>
  mutate(
    start_date = as.Date(start_date),
    end_date   = as.Date(end_date),
    days_to_election = as.integer(election_day["2012"] - end_date)
  )
```


In the first column `poll_info`, the text string contains useful information about the **year** and **state** of each poll. However, the format of this variable makes filtering by either of these factors tedious. So, we should extract the year and state, and create a variable column for each. We do this for both the 2012 and 2016 datasets.

Although it may seem redundant to include the year as a separate variable now, when the datasets are combined later on, it will be useful to already have the distinction between the two be made.

```{r}
#| label: clean-data-extract-year-state

# Extract the year and state name for each row
polls_2016 <- polls_2016 |>
  mutate(
    year = str_extract(poll_info, "^\\d{4}"),
    state = str_extract(poll_info, "(?<=^\\d{4}-).*?(?=-president)")
  )

polls_2012 <- polls_2012 |>
  mutate(
    year = str_extract(poll_info, "^\\d{4}"),
    state = str_extract(poll_info, "(?<=^\\d{4}-).*?(?=-president)")
  )
```

The 2012 and 2016 poll files have **different candidate columns** (Obama/Romney vs Trump/Clinton/Johnson/McMullin) and different column orders. To analyse them together, we need reshape this to a tidy long format with one row per poll per candidate. In this way, we can better compare the two datasets.

```{r}
#| label: clean-data-pivot-long
#| warning: false

# Create longer datasets with each candidate on a new row
candidate_cols_2016 <- c("Trump", "Clinton", "Johnson", "McMullin", 
                         "Other", "Undecided")
polls_2016_long <- polls_2016 |>
  pivot_longer(cols = all_of(candidate_cols_2016),
               values_transform = as.numeric,
               names_to = "candidate", values_to = "pct")

candidate_cols_2012 <- c("Obama", "Romney", "Other", "Undecided")
polls_2012_long <- polls_2012 |>
  pivot_longer(cols = all_of(candidate_cols_2012),
               values_transform = as.numeric,
               names_to = "candidate", values_to = "pct")
```


From inspection, there are some `sample_size` values entered as "-1". These should be NAs, rather than being treated as valid data points as they skew the mean slightly toward zero. First, we need to check how many rows are affected.

```{r}
#| label: clean-data-sample-size-nas

polls_2012_long |>
  filter(sample_size == -1) |> count()

polls_2016_long |>
  filter(sample_size == -1) |> count()
```


There are 4 rows in `polls_2012_long` and 6 in `polls_2016_long` with a negative sample size, which we converted to `NA`s.

We also converted the variables `sample_subpopulation`, `mode`, `partisanship`, `partisan_affiliation`, `state`, and `candidate` to **factors rather than characters** so that they can be used more easily, mainly for plotting. We then combined the two datasets into one dataset containing opinion polling results from both 2012 and 2016. Finally, all datasets were saved into the clean data folder.

```{r}
#| label: clean-data-combine-long

# Combine datasets into long set and remove negative population values
polls_long <- bind_rows(polls_2012_long, polls_2016_long) |>
  mutate(sample_size = na_if(sample_size, -1),
         across(c(sample_subpopulation, mode, partisanship,
                  partisan_affiliation, state, candidate),
                factor)
  )
```

```{r}
#| label: save-data-clean
#| echo: FALSE

write_csv(polls_long, "data/clean/polls_long_clean.csv")    # combined date
write_csv(polls_2012, "data/clean/polls_2012_clean.csv")    # clean 2012 data
write_csv(polls_2016, "data/clean/polls_2016_clean.csv")    # clean 2016 data
write_csv(polls_2012_long, "data/clean/polls_2012_long_clean.csv")  # 2012 long
write_csv(polls_2016_long, "data/clean/polls_2016_long_clean.csv")  # 2016 long
```

We also need to clean `results_1976_2024_raw`. There are several pointless variables, `state_fips`, `state_ic`, `state_cen`,`state_po`, `notes`, and `office`. The latter four are mainly for state identification purposes and not useful for statistical analysis purposes at all. The variables `notes` and `office` are also useless as they either contain all NA's or the same value for all observations. Then, the names each level of the categorical variables (`state`, `party_simplified`, `party_detailed`) are first cleaned using `str_to_kebab` before being converted to factors. Additionally, we create the variable `percent_voted` to replace `candidate_votes` by dividing `candidate_votes` by `total_votes` as it is more useful for analysis and also saves us from calculating the percentage every time we need it.

The process for cleaning the `version` variable is significantly more involved. While it is intended to contain dates, it contains a mix of strings of the form "YYYYMMDD" and "DD/MM/YYYY". To convert everything to a date of form YMD, we must pass strings of form "YYYYMMDD" through `ymd()`. On the other hand, we must also convert strings of form "DD/MM/YYYY" to "DD-MM-YYYY", then pass them through `dmy()`. Note that this is necessary as "DD/MM/YYYY" is not a valid format for the `dmy()` function. To implement this, we start by defining the temporary boolean variable `has_slash` which is only true when a slash is detected in a string. This creates two new columns which  then become the new `version` variable. The first is defined as follows: when `has_slash` is true, slashes are replaced with "-" before being passed through `dmy()`, and converted to an NA otherwise. As for the second: when `has_slash` is true, it is converted to NA, and passed through `ymd()` otherwise. We then combine these into one using `coalesce()` which only keeps non-NA values from both columns. Finally, missing all rows with missing values are omitted.

```{r}
#| label: clean-results-1976-2024

results_1976_2024_clean <- results_1976_2024_raw |>
  select(-c(state_fips, state_ic, state_cen, notes, office, state_po)) |>
  mutate(
    across(
      c(state, party_simplified, party_detailed),
      ~.x |> str_to_kebab() |> factor()
      ),
    percent_voted = round((candidatevotes / totalvotes) * 100, 2),
    has_slash = str_detect(version, "/"), #temporary column used to convert version to date
    version = coalesce(
      dmy(str_replace_all(if_else(has_slash, version, NA_character_), "/", "-")),
      ymd(if_else(has_slash, NA_character_, version))
      ),
    state = state |> fct_recode("washington-dc" = "district-of-columbia")
  ) |>
  select(-c(has_slash, totalvotes)) |>
  na.omit()
```

The socioeconomic datasets gathered externally contain mean family income, percentage of families living below the poverty line, percentage of unemployment, percentage of population covered by 100% smokefree bar laws, and the percentage of population with less than a 9th grade education, all by state. We use these datasets as they could provide valuable insights when analysed together with the polling data.

Upon inspection, they all contain the `fips` variable which represented the state code. We remove this as it is not useful to us. Additionally, they all contain a row for the United States as a whole and Puerto Rico, which are both not included in the other datasets and we also remove accordingly. Lastly, we convert the `state` variable to a factor where the level "district-of-columbia" is recoded as "washington-dc" to be consistent with the other datasets. As all datasets share these problems, we create the `fast_clean()` function to simplify the cleaning code.

```{r}
#| label: socioeconomic-data-clean

fast_clean <- function(raw_dataset)
{
  new_dataset <- raw_dataset |> clean_names() |>
    filter(!(state %in% c("Puerto Rico", "United States"))) |>
    select(-fips) |>
    arrange(state) |>
    mutate(
      state = factor(str_to_kebab(state)) |> 
        fct_recode("washington-dc" = "district-of-columbia")
    ) |>
    na.omit()
  return(new_dataset)
}
```

We use this function to clean all datasets, where variables are renamed accordingly to be more descriptive, and also remove any redundant variables.

```{r}
#| label: socioeconomic-data-clean-continued

income <- income_raw |> fast_clean() |>
  rename(family_income = dollars) |>
  select(-3)

poverty <- poverty_raw |> fast_clean() |>
  rename(percent_families_below_poverty = percent) |>
  select(-c(3, 4))

unemployment <- unemployment_raw |> fast_clean() |>
  rename(percent_unemployed = percent) |>
  select(-c(3, 4))

smokefree_rate <- smokefree_rate_raw |> fast_clean() |>
  rename(percent_smokefree = percent_1)

education <- education_raw |> fast_clean() |>
  rename(percent_uneducated = percent) |>
  select(-people_education_less_than_9th_grade, -c(3, 4))
```

Finally, we combine all datasets into one, called `socioeconomic_clean` and save it in to the clean data folder.

```{r}
#| label: socioeconomic-combine-datasets

df_list <- list(income, poverty, unemployment, smokefree_rate, education)

socioeconomic_clean <- df_list |> reduce(left_join, by = "state")
write_csv(socioeconomic_clean, glue("{clean_dir}/socioeconomic.csv")) #Socioeconomic data
```


- What does this data represent?
- What is measured?
- How was it collected?

*Is there any data missingness or other errors? How can this be corrected?*

- Do the measurements make sense in context? (you might need to discuss with the instructor what to do if not)




# Exploratory Data Analysis

(Use summaries and visualisations to identify relationships in the data, and motivate your research questions)

*What relationships can you observe visually?*
Here, maybe include our graphs from week 3 forum?

```{r}
#| label: Count of Polls per State
#| echo: FALSE

polls_2016 |> count(state, sort = TRUE) |> print(n = Inf)

polls_2012 |> count(state, sort = TRUE) |> print(n = Inf)
```
Discussion: In 2016, 51 states were included and in 2012, 48 states were included in the pre-polling data. There was also a large range of the number of polls per state. For example, in 2012 there were 110 polls from Ohio compared to the single poll from Alabama. It means that any nation wide statement would be more biased towards the states that had more polls.


```{r}
#| label: Candidates 2012 v Percentage (pre-polling)
#| echo: FALSE

polls_2012_long |> filter(year %in% c("2012")) |>
  ggplot() + 
  geom_boxplot(aes(x = candidate, y = pct, fill = candidate)) +
  labs(x = "Candidates 2012", y = "Percentage") + 
  theme_bw()
```

Discussion: From this data set we could explore how accurate the pre-polling data was to the outcome data from the results data set. This may need to be changed because the American voting system depends on the state and each state had different amount of polls which means states that has less polls or was not included will not be represented. This will bias the pre-polling towards highly polled states.

```{r}
#| label: plot-obama-votes-over-time
#| echo: FALSE

polls_2012 |> ggplot(aes(y = Obama,
                         x = as.integer(difftime(ymd("2012-11-06"), end_date, units = "days")))
) + 
  geom_point(color = "darkcyan", shape = 1, stroke = 1, alpha = 0.6) +
  labs(x = "Time until election (days)",
       y = "Percent voted for Obama",
       title = str_wrap("Relationship between percentage of people 
       who voted for Obama and time until election day")) +
  scale_x_reverse() +
  scale_y_continuous(limits = c(0, NA))
```

Discussion: The purpose of this graph to to see how pre-polling votes change over time for a Obama in 2012. Just by looking at the data there does not seem to be much change in opinions over time. The main aspect to see is that more polls were completed closer to the election date. We may be able to find a better trend with more analysis. Exploring voting opinions over time may be limited by the lack of data for more than 300 days before election. Depending what we are trying to explore we may not have not enough data.

```{r}
#| label: time-plot
#| echo: FALSE

#| fig-cap: "Temporal distribution of polling effort, 2016."
polls_2016 |>
  ggplot(aes(days_to_election)) +
  geom_histogram(binwidth = 7) +
  scale_x_reverse() +
  labs(x = "Days to election (binned by week)", y = "Number of polls")

#| label: time-plot
#| fig-cap: "Temporal distribution of polling effort, 2012."
polls_2012 |>
  ggplot(aes(days_to_election)) +
  geom_histogram(binwidth = 7) +
  scale_x_reverse() +
  labs(x = "Days to election (binned by week)", y = "Number of polls")
```

Discussion: Histogram of days to election versus number of polls taken. More polls were undertaken closer to the election day. This graph shows very similar data as the Relationship between percentage of people who voted for Obama and time until election day

```{r}
#| label: plot-partisan-affilitation-votes-for-obama
#| echo: FALSE

polls_2012_long |> filter(candidate %in% c('Obama')) |>
  ggplot() +
  geom_boxplot(aes(x = partisan_affiliation, y = pct, fill = partisan_affiliation)) +
  labs(x = "Partisan Affilitation", y = "Percentage of Poll for Obama") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_fill_viridis_d()
```

Discussion: This graph focuses on the votes for a specific candidate in a election for each of the partisan affiliations. This will give incite into whether participation affiliation of the pollster will influence the percentage of the votes for each party. This graph is interesting because there are differences in the median vote percentages but also variety in the size of the interquartile ranges. This may be caused by sample size difference though because none partisan affiliation is the most common type and has the largest interquartile range. This may also be influenced by location as where the pollster are located may effect votes.

```{r}
#| label: plot-voting-methods-for-romney
#| echo: FALSE

ggplot(polls_2012, aes(x = mode, y = Romney, fill = mode)) +
  geom_boxplot() +
  labs(x = "Method of Voting", y = "Percentage of Votes for Romney", title = "Percentage of votes for romney in each state versus the method of pre-polling, for the 2012 american presidential election") +
  theme_bw() +
  theme(legend.position = "none")
```

Discussion: The method of voting influences the percentage of votes for a candidate as the median changes with the method of voting and along with the interquartile range. There is also overlap in the methods as mixed could include multiple. To improve this graph Live Phone/Online could be included in mixed as there looks to be only 1 data point. It will also need to be considered the different demographics for each methods. For example the demographic that will answer by mail will be different from the ones through the internet. This would mean that the correlation may be within demographics not the methods of voting. 

```{r}
#| label: plot-results-2016-percentage-of-state-votes
#| echo: FALSE

results_1976_2024_clean |> filter(year %in% (2016)) |>
  ggplot(aes(x = party_simplified, y = (percent_voted), fill = party_simplified)) +
  geom_boxplot() +
  labs(x = "Party", y = "Percentage of Votes in a State", title = "Histogram of percentage of votes per party in each state, in the 2016 american presidential election ") +
  theme_bw() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))
```

Discussion: Histogram of percentage of votes in each state for each party. Graph indicates most votes going towards Democrat and Republican candidates. The biggest limitation of this graph is if it was to be used to calculate the percentages of votes for a candidate overall will be different to the American voting system which gives a different weight per state by population, which is not represented in the historgram. This could be adjusted for by giving each state a weight. A similar graph can be made for other years. These results can be compared to the polling data. 


```{r}
#| label: plot-results-party-results-over-time
#| echo: FALSE

results_1976_2024_clean |>
  ggplot(aes(x = as.numeric(year), y = (candidatevotes/1000000), fill = party_simplified)) +
  geom_col() +
  labs(x = "Year", y = "Number of votes (millions)", fill = "Party", title = "Number of votes for each party over time") +
    scale_y_continuous() +
  theme_bw() +
  theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5))
```

Discussion: Column graph of votes over time. Democrat and republican parties got majority of the votes. Trend of more people voting over time probably due to population increase. Does not appear to be an trend over time of what parties people vote for. This graph has similar limitations about the weighting of states. 

```{r}
#| label: col-graph-results-2016-percentage-votesin each state
#| echo: FALSE

results_1976_2024_clean |> filter(year %in% (2016)) |> 
  ggplot(aes(x = state, y = percent_voted, fill = party_simplified)) +
  geom_col() +
  labs(x = "State", y = "Percentage of Votes (%)", fill = "Party", title = "Percentage of votes per party in each state") +
 # scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  theme_bw() +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), plot.title = element_text(hjust = 0.5)) 
```

Discussion: Column graph of percentage of votes by state. Democrat and republican parties have majority of the votes. This way of representing the data is less limited as it make more sense to look at data by state because of the way the American election system works. The District of Columbia stands out as the large majority of their votes were for Democrat compared to the other states for this year.  This graph would be best to compare with the pre-polling data or to compare over time. 



# Research Questions

- How representative were opinion polls of the election results?

# Bibliography

1. 