---
title: "EDA-Report-JALT"
subtitle: "STATX290 Statistical Practice Exploratory Data Analysis Report"
author: "Team JALT - Jerry, Aysha, Lily, and Theo"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    embed-resources: false
    theme: cosmo
execute:
  warning: false
  message: false
  echo: true
---

# Introduction

This EDA report looks at polling data from 2012 and 2016 US pre-election polling. 

The data was obtained from HuffPost Pollster at https://projects.huffingtonpost.com. There are three data sets:

- `state_polls_2012.csv`
- `state_polls_2016.csv`
- `1976-2024-president.csv`

In the raw data sets, each row represents a separate poll containing the candidates for that year, the polling dates, 

An additional data set, `1976-2024-president.csv`, along with the file `codebook-us-president-1976-2020.md` is included.

# Description of Data Cleaning

- What does this data represent?
- What is measured?
- How was it collected?

*Is there any data missingness or other errors? How can this be corrected?*

- Do the measurements make sense in context? (you might need to discuss with the instructor what to do if not)

Extracting the state column
```{r}

```


# Exploratory Data Analysis

(Use summaries and visualisations to identify relationships in the data, and motivate your research questions)

*What relationships can you observe visually?*
Here, maybe include our graphs from week 3 forum?



```{r}
#| label: Candidates 2012 v Percentage (pre-polling)
#| eval: false
polls_2012_long |> filter(year %in% c("2012")) |>
  ggplot() + 
  geom_boxplot(aes(x = candidate, y = pct, fill = candidate)) +
  labs(x = "Candidates 2012", y = "Percentage") + 
  theme_bw()
```

Discussion: From this data set we could explore how accurate the pre-polling data was to the outcome data from the results data set. This may need to be changed because the American voting system depends on the state and each state had different amount of polls which means states that has less polls or was not included will not be represented. This will bias the pre-polling towards highly polled states.




# Research Questions

- How representative were opinion polls of the election results?