---
title: "Testing"
format: html
---

############################################################################
 - Maybe look into socioeconomic factors and the percentage of the population that voted
 - Does unemployment over time effect voting parties
 - it has been shown that dissatisfaction with the economy, democracy and income inequality are all positively associated with electoral support for Radical Left Parties https://www.tandfonline.com/doi/full/10.1080/14616696.2022.2127829#d1e195
 - religion effecting turnout or voting preferences
 - https://collected.jcu.edu/cgi/viewcontent.cgi?article=1042&context=honorspapers
############################################################################




```{r}
Votes_2024_3 <- filter(results_1976_2024_clean, year == "2024")
Votes_2024_merged_3 <- merge(x = Votes_2024_3, y = socioeconomic_clean, by = "state")
```


```{r}
#| label: scatter-graph-polls-2016
#| echo: FALSE

Votes_2024_merged_3 |> ggplot(aes(
  x = family_income,
  y = percent_voted,
  color = party_simplified
  )) + 
  geom_point(shape = 1, stroke = 1, alpha = 0.9) +
  labs(
    x = "Family Income",
    y = "Republican Vote percentage",
    title = "family income and party vote percentage"
  ) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, size = 11)) +
  scale_x_continuous(limits = c(75000, NA)) +
  scale_y_continuous(limits = c(0, NA))
```

```{r}
#| label: scatter-graph-polls-2016
#| echo: FALSE

Votes_2024_merged_3 |> ggplot(aes(
  x = percent_families_below_poverty,
  y = percent_voted,
  color = party_simplified
  )) + 
  geom_point(shape = 1, stroke = 1, alpha = 0.9) +
  labs(
    x = "Percent of families below poverty",
    y = "Republican Vote percentage",
    title = "Families below poverty and party vote percentage"
  ) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, size = 11)) +
  scale_x_continuous(limits = c(0, NA)) +
  scale_y_continuous(limits = c(0, NA))
```

```{r}
#| label: scatter-graph-polls-2016
#| echo: FALSE

Votes_2024_merged_3 |> ggplot(aes(
  x = percent_unemployed,
  y = percent_voted,
  color = party_simplified
  )) + 
  geom_point(shape = 1, stroke = 1, alpha = 0.9) +
  labs(
    x = "Percent Unemployed",
    y = "Republican Vote percentage",
    title = "Unemployed and party vote percentage"
  ) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, size = 11)) +
  scale_x_continuous(limits = c(0, NA)) +
  scale_y_continuous(limits = c(0, NA))
```

```{r}
#| label: scatter-graph-polls-2016
#| echo: FALSE

Votes_2024_merged_3 |> ggplot(aes(
  x = percent_smokefree,
  y = percent_voted,
  color = party_simplified
  )) + 
  geom_point(shape = 1, stroke = 1, alpha = 0.9) +
  labs(
    x = "Percent Smoke Free",
    y = "Republican Vote percentage",
    title = "Smoke Free and party vote percentage"
  ) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, size = 11)) +
  scale_x_continuous(limits = c(0, NA)) +
  scale_y_continuous(limits = c(0, NA))
```

```{r}
#| label: scatter-graph-polls-2016
#| echo: FALSE

Votes_2024_merged_3 |> ggplot(aes(
  x = percent_uneducated,
  y = percent_voted,
  color = party_simplified
  )) + 
  geom_point(shape = 1, stroke = 1, alpha = 0.9) +
  labs(
    x = "Percent Uneducated",
    y = "Republican Vote percentage",
    title = "Uneducated and party vote percentage"
  ) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, size = 11)) +
  scale_x_continuous(limits = c(0, NA)) +
  scale_y_continuous(limits = c(0, NA))
```


